Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-10-22 RMS/submission.tex
finish polishing for the moment, lacking figures demonstrate scheduling

\subsection{Real-Time Maude}
Based on \emph{rewriting logic}, \emph{Real-Time Maude} is a language
and tool that extends \emph{Maude} to support the formal specification
and analysis of real-time
systems~\cite{DBLP:journals/lisp/OlveczkyM07}.

\subsubsection{Specification}
Real-Time Maude models systems as \emph{modules}. A module specifies a
\emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup A , IR,
TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional equations on $\Sigma$, and
  $A$ a set of equational axioms such as associativity, commutativity
  and identity.  $(\Sigma, E\cup A)$ models the system's ``static''
  states as an algebraic data type, and includes a built-in
  specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instaneous transition},
  without consuming time, from an instance of $t$ to the corresponding
  instance of $t'$, \emph{provided} the conditions hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules}
  $[l]~:~\{t\}\rightarrow\{t'\} \mbox{ \textbf{in time} }r\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions} advancing
  time in the \emph{entire} state $t$ by $r$ time units. $IR$ and $TR$
  together model the ``dynamic'' behaviors of the system.
\end{itemize}

Real-Time Maude supports specifications in \emph{object-oriented}
style.  A class declaration $\texttt{class }C\texttt{ |
}att_1\texttt{:}s_1\texttt{,}\ldots\texttt{,}att_n\texttt{:}s_n$
defines a class $C$ with attributes $att_1$ to $att_n$ of sorts $s_1$
to $s_n$, respectively. An \emph{object} of class $C$ is represented
as a term $\texttt{< } O\texttt{:} C \texttt{ | }
att_1\texttt{:}val_1\texttt{,} \ldots
\texttt{,}att_n\texttt{:}val_n\texttt{ >}$ of sort \verb|Object|,
where $O$, of sort \verb|Oid|, is the object's \emph{identifier}, and
$val_1$ to $val_n$ are the values of the attributes $att_1$ to
$att_n$. A \emph{subclass} inherits all the attributes and rules of
its superclasses.

@@

In this paper, we only consider Real-Time Maude's \emph{linear
  temporal logic model checker}, which analyzes whether \emph{each}
behavior satisfies a temporal logic formula. \emph{State propositions}
are terms of sort \verb|Prop|. Their symantics is defined by equations
of the form $\texttt{ceq } statePattern \texttt{ |= } prop \texttt{ =
} b \texttt{ if } cond$, with $b$ a term of sort \verb|Bool|, stating
that $prop$ evaluates to $b$ in states which are instances of
$statePattern$ provided the condition $cond$ holds. These equations
together define $prop$ to hold in all states $t$ that make $t \texttt{
  |= } prop$ evaluate to \verb|true|. A temporal logic \emph{formula}
is constructed by state propositions and temporal logic operators such
as \verb|True|, \verb|False|, \verb|~|(negation), \verb|/\|,
\verb|\/|, \verb|->|(implication), \verb|[]|(``always''),
\verb|<>|(``eventually'') and \verb|U|(``until''). Real-Time Maude
supports both \emph{timed} and \emph{untimed} LTL model checking. The
untimed model checking command
\begin{alltt}
  (mc \(t\) |=u \(formula\) .)
\end{alltt}
checks whether the temporal logic $formula$ holds in all behaviors
starting from the initial state $t$, \emph{with no time limit}.

@@

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules. In particular, modeling of instaneous behaviors would be
explained in Sections~\ref{ss:ir}, \ref{ss:init}
and~\ref{ss:inthandling}, followed by timed behaviors in
Section~\ref{ss:timedbehavior}.

@@

We also need a sort \verb|Interval|
to denote computation requirements of tasks, 
and a sort \jx{} \verb|Counter| to
represent the time some task has been running and the corresponding
possible finishing interval.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

At last, to make our model checkable by \emph{untimed model-checking},
it is reasonable to reset the global variable $timer$ when it reaches
an upperbound while increasing (see Line~\ref{l:timer} in
Algorithm~\ref{a:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upperbound would be the least common multiple of
the periods of all tasks.
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}

\subsection{Modeling the System States}
The running system can be considered as consisting of several parts:
the tasks which are scheduled, the scheduler itself, the hardware
including registers and stacks, and the interrupt source. The
scheduler, i.e. the function $schedule()$, can be described by a
single variable $timer$. We present the models of the other parts one
by one.

\subsubsection{Tasks}
Each task is abstracted from its functionality as a \verb|Counter|.
Overhead for scheduling and switching is considered in the model. They
are treated as two system tasks. Every task is modeled as an object
instance of some subclass of the base class \verb|Task|:
\begin{verbatim}
  class Task | cnt : Counter .
  op error : -> Object [ctor] .
\end{verbatim}
The attribute \verb|cnt| denotes the computation time of the task,
while \verb|error| is an object indicating some task which misses its
deadline.

A periodic task, which is needed to be scheduled, is an object
instance of the subclass \verb|PTask| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, period : NzNat, status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, just as
in the implementation:
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
Periodic tasks are identified by their indexes in the list.

@@

\subsubsection{Hardware}
Our model considers two parts of hardware related to the interrupt
handling inside the system: the registers and the stack.

@@

\subsubsection{System}
The entire system in the model is a composition of the parts
introduced above, with a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : 
       TaskList Timer SysTasks Hardware Object ~> System [ctor] .
  mb (L T STS HW < O:Oid : IntSrc |>) : System .
\end{verbatim}

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instaneous action, and thus is modeled by the
following instaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is modeled by
the recursive operation as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and \verb|update_with_| updating the status of individual task
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule})
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, status : ST > with TIMER
        = if ST == DORMANT then < O : PTask | status : READY >
          else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > with TIMER
       = if ST == RUNNING then < O : PTask | status : INTERRUPT >
         else < O : PTask |> fi [owise] .
\end{verbatim}
In the case where \verb|TIMER|($timer$) can be divided by the task's
period, i.e. the task should be initiated, if the status is
\verb|DORMANT|, it should be ready to run; otherwise it means the
previous job of the task is not finished, thus it misses its deadline,
producing an \verb|error|. In the other case where the tasks should
not be initiated, the status of the task changes only if it is
\verb|RUNNING|. We can see that \verb|updateStatus_with_| behaves the
same as $updateStatus()$.

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the \verb|mask| flag to be cleared. Once
the request is detected, it is handled by two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on) and to call the $schedule()$
function. This is modeled by the following instaneous rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The operation \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the \verb|ir| flag, which means the request
has been handled; (ii) pushing the current \verb|pc| value into the
stack, storing the interrupted context; (iii) assigning
\verb|scheduling| of sort \verb|Oid| to \verb|pc|, which indicates
that $schedule()$ is running; and (iv) setting the \verb|mask| flag,
to mask the coming interrupt requests. As for
\verb|_.startScheduling|, it performs updating the status of
$taskList$ and increasing $timer$ by $1$, corresponding to
Lines~\ref{l:updatestatus} and \ref{l:inc} in
Algorithm~\ref{a:schedule}.
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
       = ((updateStatus L with T) inc(T) STS HW ISRC) .
\end{verbatim}

With our terminology, when the running time of the system task
\verb|scheduling| reaches its possible finishing interval, the
following rule may apply to finish the scheduling stage:
\begin{verbatim}
  crl [scheduling-finish] :
    (L T STS HW ISRC) => (SYSTEM).finishScheduling
    if SYSTEM := (L T STS HW ISRC) 
       /\ (SYSTEM).running == scheduling 
       /\ scheduling mayFinish?in STS .
\end{verbatim}
where operation \verb|_.running| returns the current \verb|pc| value
of the system and \verb|_mayFinish?in_| checks if the running time of
the task reaches its possible finishing interval:
\begin{verbatim}
  op _mayFinish?in_ : Oid SysTasks ~> Bool .
  eq O mayFinish?in [ < O : SysTask | cnt : C > REST ] 
       = C mayFinish? .
  op _mayFinish? : Counter -> Bool .
  eq [ R / [ MIN , MAX ] ] mayFinish?
       = if R lt MIN then false else true fi .
\end{verbatim}
And the operation \verb|_.finishScheduling| performs the remaining
instructions, which are left by \verb|_.startScheduling|, in the
scheduling stage: finding the first should-be-run periodic task and
setting it to run (Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}).
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
       = (L T (finish scheduling in STS) HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or letting it run, respectively.

Similar to \verb|scheduling|, the \verb|switching| stage starts when
the running time of the current running periodic task reaches its
possible finishing interval and finishes when its own running time
does so, with the two following similar rules:
\begin{verbatim}
  crl [task-finish] :
    (L T STS HW ISRC) => (SYSTEM).startSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ some N := (SYSTEM).running
       /\ some N mayFinish?in L .
  crl [switching-finish] :
    (L T STS HW ISRC) => (SYSTEM).finishSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ (SYSTEM).running == switching
       /\ switching mayFinish?in STS .
\end{verbatim}

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
\emph{Timed behaviors} of the system consist of two parts: the
execution of tasks and the execution of the interrupt source. Both are
modeled into the following one ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instaneous
transition \emph{must} be performed. Notice that the variable $R$ is
\emph{continuous} with respect to the specific time
domain\footnote{Real-Time Maude contains built-in modules to define
  the time domain to be natural numbers and rational numbers,
  specifying \emph{discrete} time domains and \emph{dense} time
  domains respectively.}  that we choose to instantiate the model on,
which is different from timed automata that discretize dense time by
defining ``clock region''.

Time affects the system by advancing both the running task whose $ID$
is loaded at \verb|pc| and the interrupt source simultaneously.  While
time elapses, \verb|cnt| of the former increases and \verb|val| of the
latter decreases, respetively.
\begin{verbatim}
  ceq delta((L T STS HW ISRC), R)
        = ((deltaTask(ID, L, R) T STS HW (deltaIS(ISRC, R)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
We omit details for the case where \verb|ID| is of sort \verb|Oid| due
to similarity. In that case, \verb|deltaTask| applies on \verb|STS|
instead of \verb|L|.

\verb|mte|, the maximum amount of time allowed to elapse, depends on
when the next instaneous action must perform. Therefore, it is decided
by three arguments: the remaining time to finish the running task, the
remaining time to request the next interrupt, and whether or not there
exists an interrupt request detected for the moment:
\begin{verbatim}
  ceq mte(L T STS HW ISRC)
        = minimum( mteTask(ID, L),
                   minimum( mteIS(ISRC), mteIr(HW)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| will return \verb|zero| if there exists a
non-masked interrupt request in the system, or \verb|INF| which
represents \emph{infinity} otherwise. Again we do not show the case
where \verb|ID| is of sort \verb|Oid|, which is very similar. We
should point out that \verb|mteTask| computes the remaining time to
reach the maximum of the possible finishing interval of the task,
since it is the time at which a complete transition \emph{must}
happen.

\section{Formal Verification}
\label{s:verification}
In this section, we analyze our model of the RMS implementation within
different realistic scenarios.  Notice that from any (reasonable)
given initial state, the number of reachable states is finite, but may
be unknown, thanks to the upper bound given to the $timer$, which
provides the potential for applying the untimed model checker.

>>>>

\subsection{Real-Time Maude}
Based on \emph{rewriting logic}~\cite{DBLP:journals/jlp/Meseguer12},
\emph{Real-Time Maude} is a language and tool that extends
\emph{Maude} to support the formal specification and analysis of
real-time systems~\cite{DBLP:journals/lisp/OlveczkyM07}.

\subsubsection{Specification}
Real-Time Maude models systems as \emph{modules}. A module specifies a
\emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup A , IR,
TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional equations on $\Sigma$, and
  $A$ a set of equational axioms such as associativity, commutativity
  and identity.  $(\Sigma, E\cup A)$ models the system's ``static''
  states as an algebraic data type, and includes a built-in
  specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instantaneous
    transition}, without consuming time, from an instance of $t$ to
  the corresponding instance of $t'$, \emph{provided} the conditions
  hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules}
  $[l]~:~\{t\}\rightarrow\{t'\} \mbox{ \textbf{in time} }r\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions} advancing
  time in the \emph{entire} state $t$ by $r$ time units. $IR$ and $TR$
  together model the ``dynamic'' behaviors of the system.
\end{itemize}

In rewriting logic, rewrite rules are applied non-deterministically,
that is, when several rules can be applied on a given $t$, any of them
may be chosen. Hence non-deterministic behaviors can be modeled
naturally in Real-Time Maude.

Real-Time Maude also supports specifications in \emph{object-oriented}
style.  A class declaration $\texttt{class }C\texttt{ |
}att_1\texttt{:}s_1\texttt{,}\ldots\texttt{,}att_n\texttt{:}s_n$
defines a class $C$ with attributes $att_1$ to $att_n$ of sorts $s_1$
to $s_n$, respectively. An \emph{object} of class $C$ is represented
as a term $\texttt{< } O\texttt{:} C \texttt{ | }
att_1\texttt{:}val_1\texttt{,} \ldots
\texttt{,}att_n\texttt{:}val_n\texttt{ >}$ of sort \verb|Object|,
where $O$, of sort \verb|Oid|, is the object's \emph{identifier}, and
$val_1$ to $val_n$ are the values of the attributes $att_1$ to
$att_n$. A \emph{subclass} inherits all the attributes and rules of
its superclasses.

@@

In this paper, we only consider Real-Time Maude's \emph{linear
  temporal logic model checker}, which analyzes whether \emph{each}
behavior satisfies a temporal logic formula. \emph{State propositions}
are terms of sort \verb|Prop|. Their symantics is defined by equations
of the form $\texttt{ceq } statePattern \texttt{ |= } prop \texttt{ =
} b \texttt{ if } cond$, with $b$ a term of sort \verb|Bool|, stating
that $prop$ evaluates to $b$ in states which are instances of
$statePattern$ provided the condition $cond$ holds. These equations
together define $prop$ to hold in all states $t$ that make $t \texttt{
  |= } prop$ evaluate to \verb|true|. A temporal logic \emph{formula}
is constructed by state propositions and temporal logic operators such
as \verb|True|, \verb|False|, \verb|~|(negation), \verb|/\|,
\verb|\/|, \verb|->|(implication), \verb|[]|(``always''),
\verb|<>|(``eventually'') and \verb|U|(``until''). Real-Time Maude
supports both \emph{timed} and \emph{untimed LTL model checking}. The
untimed model checking command
\begin{alltt}
  (mc \(t\) |=u \(formula\) .)
\end{alltt}
checks whether the temporal logic $formula$ holds in all behaviors
starting from the initial state $t$, \emph{with no time limit}.

@@

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules. In particular, modeling of instantaneous behaviors would be
explained in Sections~\ref{ss:ir}, \ref{ss:init}
and~\ref{ss:inthandling}, followed by timed behaviors in
Section~\ref{ss:timedbehavior}.

@@

We also need a sort \verb|Interval| to denote computation requirements
of tasks, and a sort \verb|Counter| to record execution of tasks.  We
call \emph{execution time} the time how long a task has been executing
for. A \verb|Counter| records the execution time and the computation
requirement of a task.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .  
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

\hide{
At last, to make our model checkable by untimed model checking, it is
reasonable to reset the global variable $timer$ when it reaches an
upperbound while increasing (see Line~\ref{l:timer} in
Algorithm~\ref{a:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upperbound would be the least common multiple of
the periods of all tasks.}

The global variable $timer$ is reset when it reaches an upper bound
while increasing, which is not shown in detail in
Algorithm~\ref{a:schedule} but is reasonable. The upper bound is the
least common multiple of the periods of all tasks. A sort \verb|Timer|
is defined:
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}

\subsection{Modeling the System States}
The system can be considered as consisting of several parts: the tasks
which are scheduled, the scheduler itself, the hardware including
registers and stacks, and the interrupt source. The scheduler,
i.e. the function $schedule()$, can be described by a single variable
$timer$. We present the models of the other parts one by one.

\subsubsection{Tasks}
Each task is abstracted from its functionality as a \verb|Counter|.
Overhead for scheduling and switching is considered in our model. They
are treated as two system tasks. Every task is modeled as an object
instance of some subclass of the base class \verb|Task|:
\begin{verbatim}
  class Task | cnt : Counter .
  op error : -> Object [ctor] .
\end{verbatim}
where \verb|error| is an object indicating some task that misses its
deadline.

A periodic task, which is needed to be scheduled, is an object
instance of the subclass \verb|PTask| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, period : NzNat, status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, same as
in the implementation:
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
Periodic tasks are identified by their indexes in the list.

@@

\subsubsection{Hardware}
Our model considers two parts of hardware related to the interrupt
handling mechanism: the registers and the stack.

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, with a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : 
       TaskList Timer SysTasks Hardware Object ~> System [ctor] .
  mb (L T STS HW < O:Oid : IntSrc |>) : System .
\end{verbatim}

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instantaneous action, thus is modeled by the
following instantaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the function \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
flag indicating there exists an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is treated as an
instantaneous action in our model. It is modeled by the recursive
function as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and function \verb|update_with_| updating the status of individual
task (Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, status : ST > with TIMER
        = if ST == DORMANT then < O : PTask | status : READY >
          else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > with TIMER
       = if ST == RUNNING then < O : PTask | status : INTERRUPT >
         else < O : PTask |> fi [owise] .
\end{verbatim}
Given a task, if \verb|TIMER|($timer$) can be divided by its period,
this task should be initiated.  In the case where the task should be
initiated, it is set to be \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

We can see that \verb|updateStatus_with_| behaves the same as
$updateStatus()$.

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the \verb|mask| flag to be cleared. Once
the request is detected, it is handled by two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the $schedule()$
function. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the \verb|ir| flag, which means the request
has been handled; (ii) pushing the current \verb|pc| value into the
stack, storing the interrupted context; (iii) assigning
\verb|scheduling| of sort \verb|Oid| to \verb|pc|, which indicates
that the system is scheduling; and (iv) setting the \verb|mask| flag,
to mask coming interrupt requests.

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Algorithm~\ref{a:schedule}. It performs updating the status of
$taskList$ and increasing $timer$ by $1$. This part is modeled by
function \verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
       = ((updateStatus L with T) inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, finding the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
       = (L T (finish scheduling in STS) HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

When the execution time of the system task \verb|scheduling| reaches
its computation requirement, the scheduling stage may finish. We model
this instantaneous action with the following rule:
\begin{verbatim}
  crl [scheduling-finish] :
    (L T STS HW ISRC) => (SYSTEM).finishScheduling
    if SYSTEM := (L T STS HW ISRC) 
       /\ (SYSTEM).running == scheduling 
       /\ scheduling mayFinish?in STS .
\end{verbatim}
where function \verb|_.running| returns the current \verb|pc| value of
the system, and \verb|_mayFinish?in_| checks whether the execution
time of the task reaches its computation requirement.
\hide{
\begin{verbatim}
  op _mayFinish?in_ : Oid SysTasks ~> Bool .
  eq O mayFinish?in [ < O : SysTask | cnt : C > REST ] 
       = C mayFinish? .
  op _mayFinish? : Counter -> Bool .
  eq [ R / [ MIN , MAX ] ] mayFinish?
       = if R lt MIN then false else true fi .
\end{verbatim}}

Similar to scheduling, the switching stage is divided into timed
behaviors of \verb|switching| and its functionality. \verb|switching|
starts when the execution time of the running periodic task reaches
its computation requirement, and finishes when its own execution time
does so. Two similar instantaneous rules are defined:
\begin{verbatim}
  crl [task-finish] :
    (L T STS HW ISRC) => (SYSTEM).startSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ some N := (SYSTEM).running
       /\ some N mayFinish?in L .
  crl [switching-finish] :
    (L T STS HW ISRC) => (SYSTEM).finishSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ (SYSTEM).running == switching
       /\ switching mayFinish?in STS .
\end{verbatim}

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
\emph{Timed behaviors} of the system consist of two parts: the
execution of tasks and the execution of the interrupt source. Both are
modeled together by the following single ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instantaneous
transition \emph{must} be performed. In fact, the core to model timed
behaviors is to define functions \verb|delta| and \verb|mte|. Notice
that the variable $R$ is \emph{continuous} with respect to the
specific time domain\footnote{Real-Time Maude contains built-in
  modules to define the time domain to be natural numbers and rational
  numbers, specifying \emph{discrete} time domains and \emph{dense}
  time domains, respectively.}  that we choose to instantiate our
model on, which is different from timed automata that discretize dense
time by defining ``clock region''.

Time affects the system by advancing both the running task whose $ID$
is loaded at \verb|pc| and the interrupt source simultaneously.  While
time elapses, \verb|cnt| of the former increases and \verb|val| of the
latter decreases, respectively:
\begin{verbatim}
  ceq delta((L T STS HW ISRC), R)
        = ((deltaTask(ID, L, R) T STS HW (deltaIS(ISRC, R)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
We omit details for the case where \verb|ID| is of sort \verb|Oid| due
to similarity. In that case, \verb|deltaTask| applies on \verb|STS|
instead of \verb|L|.

\verb|mte|, the maximum amount of time allowed to elapse, depends on
when the next instantaneous action must perform. Therefore, it is
decided by three arguments: the remaining time to finish the running
task, the remaining time to request the next interrupt, and whether or
not there exists an interrupt request detected for the moment:
\begin{verbatim}
  ceq mte(L T STS HW ISRC)
        = minimum( mteTask(ID, L),
                   minimum( mteIS(ISRC), mteIr(HW)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| returns \verb|zero| if there exists an interrupt
request detected in the system, or \verb|INF| which represents
\emph{infinity} otherwise. Again we do not show the case where
\verb|ID| is of sort \verb|Oid|, which is very similar. We should
point out that \verb|mteTask| computes the remaining time to reach the
maximum of the computation requirement of the task, since that is the
time at which a \verb|task-finish| transition \emph{must} happen.

\section{Formal Verification}
\label{s:verification}
In this section, we analyze our model of the RMS implementation within
different realistic scenarios.  Notice that from any (reasonable)
given initial state, the number of reachable states is finite, but may
be unknown, thanks to the upper bound given to $timer$, which provides
the potential for applying the untimed model checker.
